---
title: "rtestim Package Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rtestim Package Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, collapse=TRUE}
library(rtestim)
library(ggplot2)
theme_set(theme_bw())
```

\newcommand{\argmin}{\mathop{\mathrm{argmin}}}
\newcommand{\Rtseq}{\mathop{\{R_t\}_{t=1}^n}}

```{r, include = FALSE}
knitr::opts_chunk$set(
  dpi = 300, 
  collapse = FALSE, 
  comment = "#>",
  fig.asp = 0.618,
  fig.width = 6,
  out.width = "80%"
)
```


# Overview

This package uses Poisson likelihood with 
trend filtering penalty (a type of regularized nonparametric regression)
to estimate the effective reproductive number, $R_t$. 
This value roughly says "how many new infections will result from
each new infection today". Values larger than 1 indicate that an 
epidemic is growing while those less than 1 indicate decline.

This vignette provides a few examples to demonstrate the usage of `{rtestim}` to 
estimate the effective reproduction number, $R_t$. `{rtestim}` 
finds a sequence of $R_t$, $\Rtseq$ of an infectious disease by solving 
the following penalized Poisson regression
\begin{equation} \label{eq:objective_fn}
    \hat{\theta} = \argmin_{\theta} \frac{1}{n} \sum_{t=1}^n \left(e^{\theta_{t}}x_t - y_t\theta_{t}\right) + \lambda \Vert D^{(k+1)} \theta \Vert_1
\end{equation}
where $y_t$ is the an epidemic signal, ideally, incident infections, but most
frequently, incident cases, on day $t$, $\theta_{t} = \log(R_t)$ is the 
natural logarithm of $R_t$ at time $t$, $D^{(k)}$ is the $k$-th order 
divided difference operator ($k \geq 0$).
The penalty $\Vert D^{(k+1)} \theta \Vert_1$ imposes smoothness on the solution
and $\lambda$ controls the level of this smoothness, with larger $\lambda$
resulting in smoother estimates.

In particular
\begin{equation}
x_t = \sum_{a = 1}^m y_{t-a} w_a
\end{equation}
is the weighted sum of previous incidence at $t$, calculated by convolving the
preceding $m$ days of new infections with the discretized serial interval distribution $w$ of length $m$. This delay distribution encapsulates the duration
of time that a previous infection is likely to lead to future infection.

To compute $\Rtseq$ with `{rtestim}`, the minimal information needed is the 
new case counts at days up until $t$ and a parametric form for the serial 
interval distribution (a Gamma density). By default, 
$2.5$ is used for both the scale and shape parameters, based on the literature
on contract tracing, representing the typical delay between case onsets.
This is discretized to be supported on the integers. 
The order of the difference operator, the degree of smoothness, defaults
to $3$. 
The sequence of smoothness penalty $\lambda$, if no $\lambda$ is provided, 
is calculated internally by the algorithm.

# Example - synthetic dataset

## Quick start

We first demonstrate the usage of the package on synthetic data, where the
new daily case counts are generated from a Poisson distribution with mean
parameter that roughly follows a wave. Note that the first observation must 
be strictly larger than 0.

```{r, fig.align='center'}
set.seed(1234)
case_counts <- c(1, rpois(100, dnorm(1:100, 50, 15) * 500 + 1))
ggplot(data.frame(x = 1:101, case_counts), aes(x, case_counts)) +
  geom_point(colour = "cornflowerblue") +
  labs(x = "Time", y = "Case Counts")
```

Next, we fit the model and visualize the resulting $\Rtseq$:

```{r, fig.align='center'}
mod <- estimate_rt(observed_counts = case_counts, nsol = 20)
plot(mod) + geom_hline(yintercept = 1)
```

`{rtestim}` estimates a spectrum of $\Rtseq$s for a range of $\lambda$ values,
where each $\Rtseq$ corresponds to a specific $\lambda$ value. If no $\lambda$ 
value is supplied by the user, `{rtestim}` will automatically calculate a sequence 
of $\lambda$ values. The additional parameter `nsol = 20` specifies the number of 
$\lambda$s for which the $\Rtseq$ is calculated


## Cross Validation

`{rtestim}` also provides a cross validation procedure for selecting the amount of smoothness to be used in the final estimate (leave-every-k-th-out cross validation). Minimizing this metric, in principle, balances prediction error
and smoothness (`lambda.min`) though if smoother estimates are desired, one
can instead use `lambda.1se`, the largest value of $\lambda$ within one standard
error of the minimum.

```{r}
mod_cv <- cv_estimate_rt(observed_counts = case_counts)
```

The following command plots the cross validation errors for each $\lambda$ in
ascending order.

```{r, fig.align='center'}
plot(mod_cv)
```

The plot above displays vertical lines that correspond to the cross-validation scores 
for specific values of $\lambda$. The blue point at the center of each line 
represents the mean score for that value of $\lambda$ across all cross-validation
folds. The top and bottom caps of each line indicate one cross-validation 
standard error above and below the mean score for the given value of $\lambda$ 
across all cross-validation folds. Two special values of $\lambda$'s are 
highlighted with dashed lines. The one on the left represents the $\lambda$ that
gives minimum mean cross-validated error, called `lambda.min`, and the one on the
right gives the most regularized model such that the cross-validated error is 
within one standard error of the minimum, called `lambda.1se`.

Users may wish to visualize the particular $\Rtseq$ which minimizes the 
cross-validation error while prioritizing smoothness.

```{r, fig.align='center'}
plot(mod_cv, which_lambda = 'lambda.1se')
```


## Uneven Reporting Frequency

Ideally, case counts are observed at regular intervals, such as daily or 
weekly, but this is not always the case. 
`{rtestim}` also accommodates scenarios in which cases are reported 
with uneven intervals. To demonstrate this, we generate a sequence of integers
representing the days at which we observe the case counts.

```{r}
observation_incr <- rpois(101, lambda = 1)
observation_incr[observation_incr == 0] = 1
observation_time <- cumsum(observation_incr)
```

We can then fit the model by passing the observation time point as `x`.

```{r, fig.align='center'}
mod <- estimate_rt(observed_counts = case_counts, x = observation_time)
plot(mod)
```


## Changing degree of difference operator

The degree of the estimated penalized Poisson regression function
$k$ defaults to 3 for the algorithm, which
corresponds to a piece-wise cubic estimate $\Rtseq$. To estimate
$\Rtseq$ with piece-wise constant curves for example, use the command

```{r, fig.align='center'}
mod <- estimate_rt(observed_counts = case_counts, korder = 0, nsol = 20)
plot(mod)
```



# Example - Canadian Covid-19 cases

Finally, we use a long history of real case counts in Canada. The data is
available from [opencovid.ca](https://opencovid.ca) and the version downloaded
on 4 July 2023 is included in the package. We use this data to estimate $R_t$.

```{r, warning=FALSE, fig.align='center'}
can <- estimate_rt(
  observed_counts = cancovid$incident_cases, 
  x = cancovid$date, 
  nsol = 20, 
  maxiter = 1e6
)

plot(can) + 
  geom_hline(yintercept = 1) +
  coord_cartesian(ylim = c(0.5, 2)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")
```



